---
title: "SPRIS Proj 3 Analysis"
author: "Gustavo Garcia-Franceschini"
date: "`r Sys.Date()`"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(gtsummary)
library(lme4)
```

# Loading data

```{r}
baseline = read_csv("baseline.csv")
endpoints = read_csv("endpoints.csv")
```

# Creating the 'sequence' variable

```{r}
baseline = baseline %>%
  mutate(sequence = case_when(
    
    period1 == "Pill A" & period2 == "Gel B" ~ 0,
    
    period1 == "Gel B" & period2 == "Pill A" ~ 0,
    
    period1 == "Pill A" & period2 == "Gel C" ~ 1,
    
    period1 == "Gel C" & period2 == "Pill A" ~ 1,
    
    period1 == "Gel C" & period2 == "Gel B" ~ 2,
    
    period1 == "Gel B" & period2 == "Gel C" ~ 2,
    
  
  ))
```


# Table 1

```{r}
baseline %>%
  mutate(gender = if_else(gender == 0, "Male", "Female")) %>%
  select(c(age, race, gender, sequence, bviral0, sviral0,
           period1, period2, period3)) %>%  
  tbl_summary(
    by = sequence,
    statistic = list(
            all_continuous() ~ "{mean} ({sd})"
        )) %>%
  add_p() %>%
  modify_caption("**Patient Characteristics**") 
```

# Tidying data

```{r}
base_df = baseline |> 
  mutate(
    ptid = factor(ptid),
    seq = paste0(
      substr(period1, nchar(period1), nchar(period1)),
      substr(period2, nchar(period2), nchar(period2)),
      substr(period3, nchar(period3), nchar(period3))),
    seq2 = case_when(
      period3=="Gel C" ~ 0,
      period3=="Gel B" ~ 1,
      period3=="Pill A" ~ 2),
    b1 = bviral1-bviral0,
    b2 = bviral3-bviral2,
    b3 = bviral5-bviral4,
    b4 = bviral6,
    s1 = sviral1-sviral0,
    s2 = sviral3-sviral2,
    s3 = sviral5-sviral4,
    s4 = sviral6) |> 
    select(ptid, seq, seq2, age, race, gender, b1, b2, b3, b4,
           s1, s2, s3, s4) |> 
    pivot_longer(b1:s4,
                 names_to = "index",
                 values_to = "viral") |> 
    mutate(period=substr(index,2,2),
           index = ifelse(substr(index,1,1)=="b","bviral","sviral")) |> 
  pivot_wider(names_from = index,
              values_from = viral)

end_df = 
  endpoints |> mutate(
    ptid = factor(ptid),
    seq = paste0(
      substr(period1, nchar(period1), nchar(period1)),
      substr(period2, nchar(period2), nchar(period2)),
      substr(period3, nchar(period3), nchar(period3))),
    seq2 = case_when(
      period3=="Gel C" ~ 0,
      period3=="Gel B" ~ 1,
      period3=="Pill A" ~ 2)) |> 
  pivot_longer(AE_pillA_week1:Adhere_gelC_week4,
               names_to = "index",
               values_to = "value") |> 
  select(ptid, seq, seq2, index, value) |> 
  mutate(week = substr(index,nchar(index),nchar(index)),
         treatment = str_sub(index, -7, -7),
         period =str_locate(seq, str_sub(index, -7, -7))[, "start"],
         index = str_sub(index,1,2)) |> 
  pivot_wider(names_from = index,
              values_from = value)

df <- 
  merge(end_df, base_df, by = c("ptid","seq","seq2","period")) |> 
  mutate(
    ptid= factor(ptid),
    period = factor(period),
    week = factor(week),
    seq  = factor(seq),
    seq2 = factor(seq2)
  )

```

```{r}
pri_df <- df %>%
  group_by(ptid, period, treatment) %>%
  summarise(
    age = first(age),
    race = first(race),
    gender = first(gender),
    seq = first(seq),
    seq2 = first(seq2),
    AE = sum(AE),
    AD = sum(Ad),
    bviral = first(bviral),
    sviral = first(sviral)
  )
```

# Have to check for remaining effect

```{r}
cross_df = pri_df %>%
  group_by(ptid) %>%
  summarize(
    seq2 = first(seq2),
    AD = sum(AD),
    AE = sum(AE),
    bviral = sum(bviral), #careful abt these interpretations: the sum of changes
    sviral = sum(sviral)
  )
```

## Chi-Square test for Homogeneity

### For Adverse events

```{r}
tab = cross_df %>%
  mutate(AE = if_else(AE == 0 , 0, 1)) %>%
  select(seq2, AE) %>%
  table()

a = tab[c(1,2), ]

b = tab[c(1,3), ]

c = tab[c(2,3), ]


test_1_2 <- chisq.test(a)
test_1_3 <- chisq.test(b)
test_2_3 <- chisq.test(c)

# Collect the p-values
p_values <- c(test_1_2$p.value, test_1_3$p.value, test_2_3$p.value)

# Apply the Bonferroni correction
bonferroni_p_values <- p.adjust(p_values, method = "bonferroni")

print("Bonferroni-adjusted p-values:")
print(bonferroni_p_values)
```



## Wilcoxon rank test

### For Adherence

```{r}
a= cross_df %>%
  select(seq2, AD) %>%
  filter(seq2 != 2) %>%
  wilcox.test(AD ~ seq2, data = .,
        exact = FALSE)

a = a$p.val

b= cross_df %>%
  select(seq2, AD) %>%
  filter(seq2 != 1) %>%
  wilcox.test(AD ~ seq2, data = .,
        exact = FALSE)

b = b$p.val

c = cross_df %>%
  select(seq2, AD) %>%
  filter(seq2 != 0) %>%
  wilcox.test(AD ~ seq2, data = .,
        exact = FALSE)

c = c$p.val

p.adjust(c(a, b, c), method = "bonferroni")
```

Result: No test rejected the nUll Hypothesis, despite Bonferroni correction

### For Adverse Events

#### Seq 0 vs Seq 1

```{r}
a= cross_df %>%
  select(seq2, AE) %>%
  filter(seq2 != 2) %>%
  wilcox.test(AE ~ seq2, data = .,
        exact = FALSE)

a = a$p.val

b = cross_df %>%
  select(seq2, AE) %>%
  filter(seq2 != 1) %>%
  wilcox.test(AE ~ seq2, data = .,
        exact = FALSE)

b = b$p.val

c = cross_df %>%
  select(seq2, AE) %>%
  filter(seq2 != 0) %>%
  wilcox.test(AE ~ seq2, data = .,
        exact = FALSE)

c = c$p.val

p.adjust(c(a, b, c), method = "bonferroni")
```

Result: No test rejected the Null Hypothesis, despite Bonferroni correction.

### For blood viral loads

```{r}
a = cross_df %>%
  select(seq2, bviral) %>%
  filter(seq2 != 2) %>%
  wilcox.test(bviral ~ seq2, data = .,
        exact = FALSE)

a = a$p.val

b = cross_df %>%
  select(seq2, bviral) %>%
  filter(seq2 != 1) %>%
  wilcox.test(bviral ~ seq2, data = .,
        exact = FALSE)

b = b$p.val

c = cross_df %>%
  select(seq2, bviral) %>%
  filter(seq2 != 0) %>%
  wilcox.test(bviral ~ seq2, data = .,
        exact = FALSE)

c = c$p.val 

p.adjust(c(a, b, c), method = "bonferroni")
```

Result: No test rejected the Null Hypothesis, despite Bonferroni correction

### For skin viral loads

```{r}
a = cross_df %>%
  select(seq2, sviral) %>%
  filter(seq2 != 2) %>%
  wilcox.test(sviral ~ seq2, data = .,
        exact = FALSE)

a = a$p.val

b = cross_df %>%
  select(seq2, sviral) %>%
  filter(seq2 != 1) %>%
  wilcox.test(sviral ~ seq2, data = .,
        exact = FALSE)

b = b$p.val

c = cross_df %>%
  select(seq2, sviral) %>%
  filter(seq2 != 0) %>%
  wilcox.test(sviral ~ seq2, data = .,
        exact = FALSE)

c = c$p.val

p.adjust(c(a, b, c), method = "bonferroni")
```

Result: No test rejected the Null Hypothesis, despite Bonferroni correction

# Primary Objectives

1) whether the medications are safe for patients,

```{r}
model = lmer(AE ~ period + treatment + age + race + gender + 
       (1 | ptid), 
     data = pri_df, 
     REML = FALSE)  # ML estimation

model = glmer(AE ~ period + treatment + age + race + gender + 
       (1 | ptid),
family = 'binomial'(link = "logit"), data = pri_df, nAGQ = 0)

summary(model)
```


2) whether patients could easily adhere to medication schedules so that long term use is feasible.

```{r}
model = lmer(AD ~ period + treatment + age + race + gender + 
       (1 | ptid), 
     data = pri_df, 
     REML = FALSE)  # ML estimation

summary(model)
```

We possibly want to know if there is a relationship between safeness/adherence and time (so we might need time to be numeric).

# Secondary Objectives

1) Check how treatments compare in terms of viral loads in blood and skin tissues.

Blood:

```{r}
model = lmer(bviral ~ period + treatment + age + race + gender + 
       (1 | ptid), 
     data = pri_df, 
     REML = FALSE)

summary(model)
```

Skin:

```{r}
model = lmer(sviral ~ period + treatment + age + race + gender + 
       (1 | ptid), 
     data = pri_df, 
     REML = FALSE)

summary(model)
```


2) Assess the correlation of PK with adherence measures and the occurrence of adverse events.

```{r}
model = glmer(AE ~ period + age + race + gender + bviral + sviral + treatment +
       (1 | ptid),
family = 'binomial'(link = "logit"), data = pri_df, nAGQ = 0)

summary(model)
```


```{r}
model = lmer(AD ~ period + treatment + age + race + gender + bviral + sviral +
       (1 | ptid), 
     data = pri_df, 
     REML = FALSE)  # ML estimation

summary(model)
```

3) identify demographic factors associated with product adherence and whether they differ by treatment (both product and regimen)

```{r}
pri_df = pri_df %>%
  mutate(gel = if_else(treatment == "B" | treatment == "C", 1, 0),
         once_day = if_else(treatment == "B", 0, 1))

model = lmer(AD ~ age + gel + once_day  + age:gel + age:once_day +
       (1 | ptid), 
     data = pri_df, 
     REML = FALSE)  # ML estimation

summary(model)
```