---
title: "Project3 Analysis"
author: "Yuying Lu"
date: "2025-03-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = F, warning = F)
```

```{r}
library(tidyverse)
library(lme4)
library(lmerTest)
library(readxl)
library(ggplot2)
library(emmeans)
```

```{r}
baseline <- read.csv("baseline.csv")
endpoints <- read.csv("endpoints.csv")

base_df = baseline |> 
  mutate(
    ptid = factor(ptid),
    seq = paste0(
      substr(period1, nchar(period1), nchar(period1)),
      substr(period2, nchar(period2), nchar(period2)),
      substr(period3, nchar(period3), nchar(period3))),
    seq2 = case_when(
      period3=="Gel C" ~ 0,
      period3=="Gel B" ~ 1,
      period3=="Pill A" ~ 2),
    b1 = bviral1-bviral0,
    b2 = bviral3-bviral2,
    b3 = bviral5-bviral4,
    follow_b = bviral6-bviral5, 
    s1 = sviral1-sviral0,
    s2 = sviral3-sviral2,
    s3 = sviral5-sviral4,
    follow_s = sviral6 - sviral5,
    ini_b1 = bviral0,
    ini_b2 = bviral2,
    ini_b3 = bviral4,
    ini_s1 = sviral0,
    ini_s2 = sviral2,
    ini_s3 = sviral4,
    end_b1 = bviral1,
    end_b2 = bviral3,
    end_b3 = bviral5,
    end_s1 = sviral1,
    end_s2 = sviral3,
    end_s3 = sviral5) |> 
    select(ptid, seq, seq2, age, race, gender, b1, b2, b3, s1, s2, s3, 
             ini_b1, ini_b2, ini_b3, ini_s1, ini_s2, ini_s3,
           end_b1, end_b2, end_b3, end_s1, end_s2, end_s3, follow_b, follow_s) |> 
    pivot_longer(b1:end_s3,
                 names_to = "index",
                 values_to = "viral") |> 
    mutate(period=substr(index,nchar(index),nchar(index)),
           index = case_when(
             substr(index,1,1)=="b" ~ "dif_bviral",
             substr(index,1,1)=="s" ~ "dif_sviral",
             substr(index,1,5)=="ini_b" ~ "base_bviral",
             substr(index,1,5)=="ini_s" ~ "base_sviral",
             substr(index,1,5)=="end_b" ~ "end_bviral",
             substr(index,1,5)=="end_s" ~ "end_sviral"
           )) |> 
    pivot_wider(names_from = index,
              values_from = viral) 
end_df = 
  endpoints |> mutate(
    ptid = factor(ptid),
    seq = paste0(
      substr(period1, nchar(period1), nchar(period1)),
      substr(period2, nchar(period2), nchar(period2)),
      substr(period3, nchar(period3), nchar(period3))),
    seq2 = case_when(
      period3=="Gel C" ~ 0,
      period3=="Gel B" ~ 1,
      period3=="Pill A" ~ 2)) |> 
  pivot_longer(AE_pillA_week1:Adhere_gelC_week4,
               names_to = "index",
               values_to = "value") |> 
  select(ptid, seq, seq2, index, value) |> 
  mutate(week = substr(index,nchar(index),nchar(index)),
         treatment = str_sub(index, -7, -7),
         period =str_locate(seq, str_sub(index, -7, -7))[, "start"],
         index = str_sub(index,1,2)) |> 
  pivot_wider(names_from = index,
              values_from = value) |> 
  mutate(carry_over = 
           case_when(
             period == "1" ~ NA,
             period == "2" ~ substr(seq,1,1),
             period == "3" ~ substr(seq,2,2)
  ))

df <- 
  merge(end_df, base_df, by = c("ptid","seq","seq2","period")) |> 
  mutate(
    ptid= factor(ptid),
    period = factor(period),
    week = factor(week),
    seq  = factor(seq),
    seq2 = factor(seq2)
  )

df
```


- 'end_...' -- viral measured at the end of period

- 'base_...' -- viral measured at the beginning of period

- 'dif_...' = 'end_...' - 'base_...' for each period,   ... could be 'bviral'/'sviral'

include the 'base_...' in the covariates

the response could be 'end_...' or 'dif_...'



**However,** I think when testing the carry-over effect, we should focus on the 'dif_bviral' and 'dif_sviral'. But not sure



# Carry-over effect

Sum up the AE/Ad at 4 weeks within a period 

I use the total number of AE (takes value in 0, 1, 2, 3, 4) in a period, you can also try to use an indicator


```{r}
period_df <- df %>%
  group_by(ptid, period, treatment, carry_over) |> 
  summarise(
    age = first(age),
    race = first(race),
    gender = first(gender),
    seq = first(seq),
    seq2 = first(seq2),
    AE = sum(AE),  # for an indicator variable , you can use AE = ifelse(sum(AE)>=1,1,0)
    Ad = sum(Ad),
    base_bviral = mean(base_bviral),
    base_sviral = mean(base_sviral),
    dif_bviral = mean(dif_bviral),
    dif_sviral = mean(dif_sviral),
    end_bviral = mean(end_bviral),
    end_sviral = mean(end_sviral),
    .groups = "drop"
  )
period_df 
```


```{r}
summary(period_df$AE)
```



```{r}
# summation of AE, Ad across all the 3 periods
seq2_df =
  period_df |> group_by(ptid) |> 
  summarise(
    age = first(age),
    race = first(race),
    gender = first(gender),
    seq = first(seq),
    seq2 = first(seq2),
    AE_sum = sum(AE),    # for indicator, you can use AE_sum = ifelse(sum(AE)>=1,1,0)
    Ad_sum = sum(Ad),
    bv_sum = sum(bviral),
    sv_sum = sum(sviral),
    base_bv_sum = sum(base_bviral),
    base_sv_sum = sum(base_sviral),
    .groups = "drop"
  )
seq2_df

```

### Bonferroni Test

```{r}
print("Bonferroni Test for AE")
model_AE <- glm(AE_sum ~ seq2, data = seq2_df) # family = binomial
emmeans(model_AE, pairwise ~ seq2, adjust = "bonferroni")

print("Bonferroni Test for Adherence")
model_Ad <- glm(Ad_sum ~ seq2, data = seq2_df)
emmeans(model_Ad, pairwise ~ seq2, adjust = "bonferroni")

print("Bonferroni Test for Blood Viral")
model_bv <- glm(bv_sum ~ seq2, data = seq2_df)
emmeans(model_bv, pairwise ~ seq2, adjust = "bonferroni")

print("Bonferroni Test for Skin Viral")
model_sv <- glm(sv_sum ~ seq2, data = seq2_df)
emmeans(model_sv, pairwise ~ seq2, adjust = "bonferroni")
```

**Conclusion**: No significant carry-over effect under Bonferroni Test

### AVOVA Test

```{r}
print("ANOVA Test for AE")
anova_AE <- aov(AE_sum ~ seq2, data = seq2_df)
summary(anova_AE)

print("ANOVA Test for Adherence")
anova_Ad <- aov(Ad_sum ~ seq2, data = seq2_df)
summary(anova_Ad)


print("ANOVA Test for Blood Viral")
anova_bv <- aov(bv_sum ~ seq2, data = seq2_df)
summary(anova_bv)

print("ANOVA Test for Skin Viral")
anova_sv <- aov(sv_sum ~ seq2, data = seq2_df)
summary(anova_sv)

```

**Conclusion**: No significant carry-over effect under ANOVA test

### Post-hoc Test to See Which Treatments Differ

```{r}
print("Post-hoc Test for AE")
model_AE <- glm(AE_sum ~ seq2, data = seq2_df) # family = binomial
emmeans(model_AE, pairwise ~ seq2, adjust = "tukey")

print("Post-hoc Test for Adherence")
model_Ad <- glm(Ad_sum ~ seq2, data = seq2_df)
emmeans(model_Ad, pairwise ~ seq2, adjust = "tukey")

print("Post-hoc Test for Blood Viral")
model_bv <- glm(bv_sum ~ seq2, data = seq2_df)
emmeans(model_bv, pairwise ~ seq2, adjust = "tukey")

print("Post-hoc Test for Skin Viral")
model_sv <- glm(sv_sum ~ seq2, data = seq2_df)
emmeans(model_sv, pairwise ~ seq2, adjust = "tukey")
```

**Conclusion**: No significant carry-over effect under Post-hoc Test



### Comparing Model with and without carry-over variavle

Check whether the coefficient of carryover is significant.

```{r}
library(glmmTMB)

# Here I use data ` period_df`, where the AE/Ad is the summation of four weeks with a period

#### For AE #######

## with carry_over
AE_carry = glmmTMB(AE ~ treatment + period + carry_over + base_bviral + base_sviral + (1 | ptid),
                  data = period_df, family = nbinom2)
Ad_carry = glmmTMB(Ad ~ treatment + period + carry_over + base_bviral + base_sviral + (1 | ptid),
                  data = period_df, family = nbinom2)
bv_carry = lmer(dif_bviral ~ age + gender +race + treatment + period + carry_over +
                   base_bviral + base_sviral+ (1 | ptid), 
                     data = period_df)

sv_carry = lmer(dif_sviral ~ age + gender +race + treatment + period + carry_over +
                   base_bviral + base_sviral+ (1 | ptid), 
                     data = period_df)

print("Regress on AE with Carryover")
summary(AE_carry)
print("Regress on Ad with Carryover")
summary(Ad_carry)
print("Regress on Blood Viral with Carryover")
summary(bv_carry)
print("Regress on Skin Viral with Carryover")
summary(sv_carry)

```


Carryover is not significant for all the model



**Therefore,** we conclude there's no carry_over effect


# Primary Objective (Focus on AE and Ad)

```{r}
### I use data 'df', which records the AE/Ad value at each week in different periods

##### For AE #####

### include week as a continuous variable
AE_week = df |> mutate(week = as.numeric(week)) |> 
                glmmTMB(AE ~ treatment + period + week + base_bviral + base_sviral + (1 | ptid),
                              data = _, family = nbinom2)

### include week as a categorical variable
AE_week_cat = glmmTMB(AE ~ treatment + period + week + base_bviral + base_sviral + (1 | ptid),
                  data = df, family = nbinom2)

### exclude week
AE_noweek = glmmTMB(AE ~ treatment + period + base_bviral + base_sviral + (1 | ptid),
                  data = df, family = nbinom2)

print("Include Week as Continuous")
summary(AE_week)
print("Include Week as Categorical")
summary(AE_week_cat)
print("Exclude Week")
summary(AE_noweek)


# Week4 is significant, we'd better consider including week as a variable

# However, treat week as continuous variable gives better AIC


```




```{r}
### include week as a continuous variable
AE_week = df |> mutate(week = as.numeric(week)) |> 
                glmmTMB(AE ~ treatment + period + week + base_bviral + base_sviral + (1 | ptid),
                              data = _, family = poisson)

### include week as a categorical variable
AE_week_cat = glmmTMB(AE ~ treatment + period + week + base_bviral + base_sviral + (1 | ptid),
                  data = df, family = poisson)

### exclude week
AE_noweek = glmmTMB(AE ~ treatment + period + base_bviral + base_sviral + (1 | ptid),
                  data = df, family = poisson)

print("Include Week as Continuous")
summary(AE_week)
print("Include Week as Categorical")
summary(AE_week_cat)
print("Exclude Week")
summary(AE_noweek)


### same for set family = poisson

# Week4 is significant, we'd better consider including week as a variable

# However, treat week as continuous variable gives better AIC


```



```{r}
##### For Ad #####

### include week as a continuous variable
Ad_week = df |> mutate(week = as.numeric(week)) |> 
                glmmTMB(Ad ~ treatment + period + week + base_bviral + base_sviral + (1 | ptid),
                              data = _, family = nbinom2)
summary(Ad_week)
# when set family = nbinom here, the model doesn't converge with AIC = NA
# therefore I choose family = poisson


Ad_week = df |> mutate(week = as.numeric(week)) |> 
                glmmTMB(Ad ~ treatment + period + week + base_bviral + base_sviral + (1 | ptid),
                              data = _, family = poisson) 

### include week as a categorical variable
Ad_week_cat = glmmTMB(Ad ~ treatment + period + week + base_bviral + base_sviral + (1 | ptid),
                              data = df, family = poisson)

### exclude week
Ad_noweek = glmmTMB(Ad ~ treatment + period + base_bviral + base_sviral + (1 | ptid),
                  data = df, family = poisson)

print("Include Week")
summary(Ad_week)
print("Include Week categorical")
summary(Ad_week_cat)
print("Exclude Week")
summary(Ad_noweek)

### According to the AIC, we can see treat week as a continuous variable is better, with the smallest AIC

```



### For AE, Penalized Logistic

```{r}
library(glmnet)
# alpha * L1 + (1-alpha) * L2
X <- model.matrix( ~ age + gender +race + treatment + period, data = pri_df)[, -1] 
# Ridge (L2 penalty)
cv_fit <- cv.glmnet(X, pri_df$AE, family = "binomial", alpha = 0)
coef(cv_fit, s = "lambda.min")

# Lasso (L1 penalty)
AE_mod <- glmnet(X, pri_df$AE,  alpha = 1)
summary(AE_mod)

# alpha=0.5
AE_mod <- glmnet(X, pri_df$AE, alpha = 0.5)
summary(AE_mod)
```



# Secondary Objective 

### 1) 


Interested in the difference in the viral load before and after the period


#### Blood Viral

```{r}
bv_mod = lmer(dif_bviral ~ base_bviral + treatment + period + AE + Ad + (1|ptid), data = period_df, REML = F)
summary(bv_mod)
```

Only age is significant when regressing on blood viral


#### Skin Viral

```{r}
sv_mod = lmer(dif_sviral ~ base_sviral + treatment + period + AE + Ad + (1|ptid), data = period_df, REML = F)
summary(sv_mod)
```


Both AE and Ad have significant impact on bviral load and sviral load

Significant Different between B-A, B-C


For B-C

```{r}
data_new = period_df
data_new$treatment <- factor(data_new$treatment, levels = c("C","B","A"))

bv_mod = lmer(dif_bviral ~ base_bviral + treatment + period + AE + Ad + (1|ptid), data = data_new, REML = F)
summary(bv_mod)
sv_mod = lmer(dif_sviral ~ base_sviral + treatment + period + AE + Ad + (1|ptid), data = data_new, REML = F)
summary(sv_mod)


```


B-C is significnt


### If spilt week in the same period

```{r}
sec_df = end_df |> 
  pivot_longer(AE:Ad,
               names_to = 'index',
               values_to = "value") |> 
  mutate(index = paste0(index,"_week",week)) |> 
  select(-week) |> 
  pivot_wider(
    names_from = index ,
    values_from = value
  ) 
sec_df = merge(sec_df, base_df, by= c("ptid", "period","seq","seq2"))
```


```{r}
bv_sec_mod = lmer(dif_bviral ~ base_bviral + treatment + period + 
                    AE_week1+ AE_week2+ AE_week3+ AE_week4 +
                    Ad_week1+ Ad_week2+ Ad_week3+ Ad_week4 + (1|ptid), data = sec_df, REML = F)
summary(bv_sec_mod)

sv_sec_mod = lmer(dif_sviral ~ base_sviral + age + gender +race + treatment + period + 
                    AE_week1+ AE_week2+ AE_week3+ AE_week4 +
                    Ad_week1+ Ad_week2+ Ad_week3+ Ad_week4 + (1|ptid), data = sec_df, REML = F)
summary(sv_sec_mod)
```




### 2) Demographic factor assiciated with Ad


```{r}
### need to split treatment into product and regimen
dem_df = df |> 
  mutate(
    product = 
      case_when(
        treatment == "A" ~ 'pill',
        .default = 'gel'),
    regimen = 
       case_when(
        treatment == "B" ~ '3',
        .default = '1')
  )
```



#### Initial model

```{r}
Ad_mod =  dem_df |> mutate(week = as.numeric(week)) |> 
  glmmTMB(Ad ~ product + regimen + base_bviral + week +period + (1|ptid), data =_, family = poisson)
summary(Ad_mod)
```

```{r}
# adding demographic variables 
Ad_mod =  dem_df |> mutate(week = as.numeric(week)) |> 
  glmmTMB(Ad ~ product + regimen + age + gender + race + base_bviral + week +period + (1|ptid), data =_, family = poisson)
summary(Ad_mod)
```

Only age shows significant association with Adherence

#### Interaction (age with product) and (age with regimen)

```{r}
bv_mod_ap = dem_df |> mutate(week = as.numeric(week)) |> 
  glmmTMB(Ad ~ age*product + regimen +  base_bviral + week + period + (1|ptid), data =_, family = poisson)

bv_mod_ar = dem_df |> mutate(week = as.numeric(week)) |> 
  glmmTMB(Ad ~ product + age*regimen +  base_bviral + week + period + (1|ptid), data =_, family = poisson)


summary(bv_mod_ap)
summary(bv_mod_ar)

```

- the association between age and adherence has significant difference varied by product
- the association between age and adherence has significant difference varied by regimen



